Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-09-29 psy_peds_3.Rmd
data wrangling

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
---

@@

# R session information
```{r}
sessionInfo()
```

@@

# Inclusion criteria
The inclusion criteria for this study is:
- Patients inclusion criteria: 
  - Age between 11-17 at the date of visit
- Visits inclusion criteria:
  - Inpatients (at least one hospitalization > 1 day)
  - Starting date between January 1st, 2019 and July 31st, 2021
  - Comprising least one ICD-10 code pertaining to the specified ICD-10 codes list (mental health related codes)

# Variables that need to be checked/modified by each site 
Change the values of the following variables according to the specificities of your site:
- "folder_4ce_files": folder path where your phase 2.2 data files are located
- "obfuscation": determine the obfuscation threshold (FALSE if no obfuscation, or the numeric value of the obfuscation threshold if any)
- "raceAvailable": set as TRUE or FALSE depending on whether the variable is being collected at your site
- "dateFormat": specify the format of the date at your site (e.g., for "03-AUG-20", the format would be "%d-%b-%y", [see documentation](https://www.stat.berkeley.edu/~s133/dates.html))

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "../Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
```

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- "2019-01-01"
history_end_date <- "2021-05-31"
pandemic_start_date <- "2020-03-15"
start_date_plots <- "2019-03-01"
end_date_plots <- "2021-05-01"
```

# Data input
We will use as input the 2.2 data files. Specifically: 
- LocalPatientSummary
- LocalPatientObservation
- LocalPatientClinicalcourse

For sites recording the race, we will also use an additional file:
- LocalPatientRace

@@

@@

### Adding columns with hospitalization number and hospitalization length to the clinical raw file
```{r}
count_sequences_hospitalisation <- function(df, ...) {
  seq_hospitalisation_df <- data.frame(total_span = seq(min(df$days_since_admission),
                                                        max(df$days_since_admission))
  ) %>%
    left_join(df, by = c("total_span" = "days_since_admission")) %>%
    replace_na(list(in_hospital = 0))
  count_sequences <- rle(seq_hospitalisation_df$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  stopifnot(length(df$days_since_admission) == length(sequences))
  data.frame(days_since_admission = df$days_since_admission,
             n_hospitalisation = sequences,
             len_hospitalisation = sequences_len)
}
stopifnot(all(clinical_raw$in_hospital == 1))
hospitalisations_seq_df <- clinical_raw %>%
  distinct(patient_num, cohort, days_since_admission, in_hospital) %>%
  group_by(patient_num, cohort) %>%
  group_modify(count_sequences_hospitalisation)

@@

### Read the ICD10 codes file
The file containing the ICD10 psychiatric related codes is located in the `public-data` folder of the GitHub repository.

### Check that we have patients with the ICD10 mental codes
```{r}
### check that the ICD codes follow the regular expression [A-Z][0-9][0-9AB]\.?[0-9A-TV-Z]{0,4}
codesToReview <- obs_raw %>% dplyr::filter(concept_type == "DIAG-ICD10",
                                           ! grepl( "[A-Z][0-9][0-9AB]\\.?[0-9A-TV-Z]{0,4}", concept_code))
print(codesToReview)

@@

## All cohort
We are selecting encounters based on the following criteria:
- associated with at least 1 ICD codes
- with a duration of hospitalization spanning over at least two consecutive days
- Starting on or after 2019/01/01

```{r}
# Adjusting patient age at the time of visit, assuming the age used is the patients' age at the time of extraction (last discharge date available in the data)
last_data_date <- max(as.Date(demo_raw$last_discharge_date, "%Y-%m-%d"))
clinical_raw <- left_join(clinical_raw, demo_raw[c("patient_num", "cohort", "age", "sex")], by = c("patient_num", "cohort"))
clinical_raw$age_time_visit <- clinical_raw$age - floor(as.numeric(last_data_date - as.Date(clinical_raw$calendar_date, dateFormat)) / 365.25)
clinical_raw <- dplyr::select(clinical_raw, -age)

obs_filtered <- obs_raw %>% 
  filter(concept_type == 'DIAG-ICD10', 
         days_since_admission >= 0)

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))

@@

stopifnot(unique(all_codes$cohort) == "AllAdm")
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))
stopifnot(nrow(all_codes) == nrow(observation_table))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(demographic_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(observation_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(clinical_table$patient_num)))
stopifnot(nrow(distinct(all_codes, patient_num, n_hospitalisation)) == nrow(distinct(clinical_table, patient_num, n_hospitalisation)))
stopifnot(nrow(distinct(all_codes, patient_num, days_since_admission)) == nrow(distinct(observation_table, patient_num, days_since_admission)))
```

### New hospitalizations
```{r, fig.width=12, fig.height=12}

new_hospitalizations_per_day <- clinical_filtered %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(date)) %>%
  filter( in_hospital == 1, 
         date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

new_hospitalizations_per_week <- new_hospitalizations_per_day %>%
  filter(weeks >= start_date_plots) %>% 
  group_by( weeks) %>%
  summarise(count = sum(count))

new_hospitalizations_per_month <- new_hospitalizations_per_day %>%
  group_by( month) %>%
  summarise(count = sum(count))

admission_plot_count <- function(df,
                               x,
                               y, 
                               title = "New admission counts",
                               color = NULL,
                               size = NULL) {
  plot <- ggplot(df, aes_string(x = x, y = y, color = color, size = size))  +
    geom_point() +
    scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
    theme(axis.text.x=element_text(angle=60, hjust=1))+
    labs(x = "Calendar date",
         y = "Patients count", 
       title = title)
  return(plot)
}

nh_day <- admission_plot_count(df = new_hospitalizations_per_day, 
                     x = "date",
                     y = "count", 
                     title = "New admissions (per day)")

nh_week <- admission_plot_count(df = new_hospitalizations_per_week, 
                     x = "weeks",
                     y = "count", 
                     title = "New admissions (per week)")

nh_month <- admission_plot_count(df = new_hospitalizations_per_month, 
                     x = "month",
                     y = "count", 
                     title = "New admissions (per month)")

grid.arrange(nh_day,
             nh_week,
             nh_month,
             ncol=1)
```

### Total hospitalize patients
```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count = n_distinct(patient_num))

@@

#### Per week
```{r}
# counts per week and sex
patients_hospitalized_per_week_agg_sex <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( weeks, sex ) %>%
  summarise(count_sex = n_distinct(patient_num))

@@

## Second estimation: total number of patients filtering by psychiatric disorders
## Creating the inputTable
Creation of the inputTable that will be the only table used throughout the rest of the analysis
- selecting only visits with at least one ICD codes pertaining to the mental health categories
- filter visits where patients are between 11 and 17 years old
- add the demographic and hospitalization related information 
- add the ICD code description and the related disorder group

### Add the psychiatric/mental health ICD code description and filter 
```{r message=FALSE, warning=FALSE}
# Getting site name, will be used later in the analysis
site <- unique(as.character(demo_raw$siteid))

@@

### Aggregated by sex
```{r}
# counts per week and sex
patients_hospitalized_per_week_agg_sex <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( weeks, sex ) %>%
  summarise(count_sex = n_distinct(patient_num))

@@

##### Total hospitalizations
We have estimated the total number of hospitalizations in the previous steps:
- total_hospitalization_per_day
- total_hospitalization_per_week
- total_hospitalization_per_month

@@

# Per day
```{r}
thpd_psyCount <- count_icd_per_day %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = date, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by day)",
       title = "Daily patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

# Per week
```{r}
thpd_psyCount <- count_icd_per_week %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = weeks, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by week)",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

Per day
```{r}
count_disorder_group_per_day <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Per week
```{r}
count_disorder_group_per_week <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Per month
```{r}
count_disorder_group_per_month <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Months 
```{r}
# estimate the monthly counts
count_icd_per_month_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Months 
```{r}
# estimate the monthly counts
count_icd_per_month <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count_icd = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

>>>>

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

# R session information

```{r}
sessionInfo()
```

@@

# Inclusion criteria

The inclusion criteria for this study is:

-   Patients inclusion criteria:

    -   Age between 11-17 at the date of visit

-   Visits inclusion criteria:

    -   Inpatients (at least one hospitalization \> 1 day)

    -   Start date: January 1st, 2019

    -   End date: May 31st, 2021

    -   Comprising at least one ICD-10 code during hospitalization

# Variables that need to be checked/modified by each site

Change the values of the following variables according to the
specificities of your site:

1.  "folder_4ce_files": folder path where your phase 2.2 data files are
    located

2.   "obfuscation": determine the obfuscation threshold (FALSE if no
    obfuscation, or the numeric value of the obfuscation threshold if
    any)

3.   "raceAvailable": set as TRUE or FALSE depending on whether the
    variable is being collected at your site

4.   "dateFormat": specify the format of the date at your site (e.g., for
    "03-AUG-20", the format would be "%d-%b-%y", [see
    documentation](https://www.stat.berkeley.edu/~s133/dates.html))

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
data_update_date <- "2021-06-05"
```

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- "2019-01-01"
history_end_date <- "2021-05-31"
pandemic_start_date <- "2020-03-15"
start_date_plots <- "2019-01-15"
end_date_plots <- "2021-05-01"
```

# Data input

We will use as input the 2.2 data files. Specifically:

- LocalPatientSummary

- LocalPatientObservation

- LocalPatientClinicalcourse

For sites recording the race, we will also use an additional file:

- LocalPatientRace

@@

### Read the ICD10 codes file

The file containing the ICD10 psychiatric related codes is located in
the `public-data` folder of the GitHub repository.

## Data-management

### Filtering on "inpatient" status

@@

### Calculating duration of individual hospitalizations

Adding columns with hospitalization number and hospitalization length to the clinical raw file

```{r}
count_sequences_hospitalisation <- function(df, ...) {
  seq_hospitalisation_df <- data.frame(total_span = seq(min(df$days_since_admission),
                                                        max(df$days_since_admission))
  ) %>%
    left_join(df, by = c("total_span" = "days_since_admission")) %>%
    replace_na(list(in_hospital = 0))
  count_sequences <- rle(seq_hospitalisation_df$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  stopifnot(length(df$days_since_admission) == length(sequences))
  data.frame(days_since_admission = df$days_since_admission,
             n_hospitalisation = sequences,
             len_hospitalisation = sequences_len)
}
stopifnot(all(clinical_raw$in_hospital == 1))
hospitalisations_seq_df <- clinical_raw %>%
  distinct(patient_num, cohort, days_since_admission, in_hospital) %>%
  group_by(patient_num, cohort) %>%
  group_modify(count_sequences_hospitalisation)

@@

### Counting ICD10 mental health codes in the obs_raw table

```{r}
### check that the ICD codes follow the regular expression [A-Z][0-9][0-9AB]\.?[0-9A-TV-Z]{0,4}
codesToReview <- obs_raw %>% dplyr::filter(concept_type == "DIAG-ICD10",
                                           ! grepl( "[A-Z][0-9][0-9AB]\\.?[0-9A-TV-Z]{0,4}", concept_code))
print(codesToReview)

@@

## Applying inclusion criteria to filter and format the data

We are selecting hospitalizations based on the following criteria:

- Associated with at least 1 ICD code

- With a duration of hospitalization spanning over at least two consecutive days

- Starting on or after 2019/01/01

### Calculating age at hospitalization time
```{r}
# Calculating patient age at the time of visit, assuming the age used is the patients' age at the time of extraction (last discharge date available in the data)
obs_raw_age <- left_join(obs_raw, demo_raw[c("patient_num", "age")], by = c("patient_num")) %>% 
  left_join(clinical_raw[, c("patient_num", "days_since_admission", "calendar_date")], 
            by = c("patient_num", "days_since_admission"))
stopifnot(nrow(obs_raw_age) == nrow(obs_raw))
obs_raw_age$age_time_diagnosis <- obs_raw_age$age - floor(as.numeric(as.Date(data_update_date) - as.Date(obs_raw_age$calendar_date, dateFormat)) / 365.25)
```

``` {r}
obs_filtered_age <- obs_raw_age %>% 
  filter(concept_type == 'DIAG-ICD10', 
         days_since_admission >= 0)

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))
```

```{r}
# Steps necessary to filter rows:
  # - encounters without ICD
  # - ICD diagnosed outside of encounters
  # - patients from the patients table that don't have any encounters matching the criteria
clinical_table <- clinical_filtered %>% 
  inner_join( distinct(obs_filtered_age, patient_num, days_since_admission),  by = c("patient_num", "days_since_admission")) %>% 
  distinct(patient_num, n_hospitalisation) %>% 
  inner_join(clinical_filtered, by = c("patient_num", "n_hospitalisation"))
# Adding new columns
observation_table <- obs_filtered_age %>% 
  inner_join(clinical_table[, c("patient_num", "days_since_admission")], by = c("patient_num", "days_since_admission"))
demographic_table <- demo_raw %>% 
  inner_join(distinct(clinical_table, patient_num), by = "patient_num")
all_codes <- observation_table %>% 
   inner_join(clinical_table, by = c("patient_num", "days_since_admission")) %>% 
   inner_join(demographic_table[, c("patient_num", "sex")], by = "patient_num")
```

```{r}
stopifnot(unique(all_codes$cohort) == "AllAdm")
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))
stopifnot(nrow(all_codes) == nrow(observation_table))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(demographic_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(observation_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(clinical_table$patient_num)))
stopifnot(nrow(distinct(all_codes, patient_num, n_hospitalisation)) == nrow(distinct(clinical_table, patient_num, n_hospitalisation)))
stopifnot(nrow(distinct(all_codes, patient_num, days_since_admission)) == nrow(distinct(observation_table, patient_num, days_since_admission)))
```

```{r}
admission_plot_count <- function(df,
                               x,
                               y, 
                               title,
                               y_title,
                               size = NULL, 
                               color = NULL){
  plot <- ggplot(df, aes_string(x = x, y = y, color = color, size = size))  +
    geom_point() +
    scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y", limits = c(start_date_plots, end_date_plots), expand = expansion(mult = 0.01, add = 1)) +
    theme(axis.text.x=element_text(angle=60, hjust=1))+
    labs(x = "Calendar date",
         y = y_title, 
       title = title) 
  return(plot)
}

### New hospitalizations

```{r}

new_hospitalizations_per_week <- clinical_table %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(weeks)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

start_date_plots <- as.Date("2019-01-01")
end_date_plots <- as.Date("2021-05-24")

nh_week <- admission_plot_count(df = new_hospitalizations_per_week, 
                     x = "weeks",
                     y = "count", 
                     title = "New admissions (per week)", 
                     y_title = "Patient Count")

nh_week
```

### Filtering on age at hospitalization time

Keeping only patients with age between 11 and 17

```{r}

# Filtering hospitalizations where patient age is in inclusion criteria
observation_table_ado <- filter(observation_table, age_time_diagnosis >= 11 & age_time_diagnosis <= 17)

# Filtering observation tables which patient_num and days_since_admission are still in clinical table after filtering on age

clinical_test <- clinical_table[, c("patient_num", "n_hospitalisation", "days_since_admission")] %>% 
  inner_join(distinct(observation_table_ado, patient_num, days_since_admission, age_time_diagnosis), 
            by = c("patient_num", "days_since_admission"))

clinical_table_ado <- clinical_table %>% 
  inner_join(clinical_test[, c("patient_num", "n_hospitalisation", "age_time_diagnosis")],
             by = c("patient_num", "n_hospitalisation"))

observation_table_ado <- clinical_table %>% 
  inner_join(observation_table[, c("patient_num", "days_since_admission")], 
             by = c("patient_num", "days_since_admission"))

# Filtering observation tables which patient_num are still in clinical table after filtering on age
demographic_table_ado <- demographic_table %>% 
  inner_join(distinct(clinical_table_ado, patient_num), by = "patient_num")

all_codes_ado <- all_codes %>% 
   filter(age_time_visit >= 11 & age_time_visit <= 17)

head(observation_table_ado)
head(clinical_table_ado)
head(all_codes_ado)
```

```{r}
stopifnot(unique(all_codes_ado$cohort) == "AllAdm")
stopifnot(unique(all_codes_ado$in_hospital) == 1)
stopifnot(unique(all_codes_ado$len_hospitalisation > min_hosp_days))
stopifnot(nrow(all_codes_ado) == nrow(observation_table_ado))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(demographic_table_ado$patient_num)))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(observation_table_ado$patient_num)))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(clinical_table_ado$patient_num)))
stopifnot(nrow(distinct(all_codes_ado, patient_num, n_hospitalisation)) == nrow(distinct(clinical_table_ado, patient_num, n_hospitalisation)))
stopifnot(nrow(distinct(all_codes_ado, patient_num, days_since_admission)) == nrow(distinct(observation_table_ado, patient_num, days_since_admission)))
```

```{r}

@@

### Total hospitalize patients

```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count = n_distinct(patient_num))

@@

#### Per week

```{r}
# counts per week and sex
patients_hospitalized_per_week_agg_sex <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( weeks, sex ) %>%
  summarise(count_sex = n_distinct(patient_num))

@@

## Second estimation: total number of patients filtering by psychiatric disorders

## Creating the inputTable

Creation of the inputTable that will be the only table used throughout
the rest of the analysis - selecting only visits with at least one ICD
codes pertaining to the mental health categories - filter visits where
patients are between 11 and 17 years old - add the demographic and
hospitalization related information - add the ICD code description and
the related disorder group

### Add the psychiatric/mental health ICD code description and filter

```{r message=FALSE, warning=FALSE}
# Getting site name, will be used later in the analysis
site <- unique(as.character(demo_raw$siteid))

@@

### Aggregated by sex

```{r}
# counts per week and sex
patients_hospitalized_per_week_agg_sex <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( weeks, sex ) %>%
  summarise(count_sex = n_distinct(patient_num))

@@

We have estimated the total number of hospitalizations in the previous
steps: - total_hospitalization_per_day - total_hospitalization_per_week
- total_hospitalization_per_month

@@

# Per day

```{r}
thpd_psyCount <- count_icd_per_day %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = date, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by day)",
       title = "Daily patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

# Per week

```{r}
thpd_psyCount <- count_icd_per_week %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = weeks, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by week)",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

Per day

```{r}
count_disorder_group_per_day <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Per week

```{r}
count_disorder_group_per_week <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Per month

```{r}
count_disorder_group_per_month <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Months

```{r}
# estimate the monthly counts
count_icd_per_month_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

@@

Months

```{r}
# estimate the monthly counts
count_icd_per_month <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count_icd = n_distinct(patient_num)) %>%
  mutate( period = ifelse( month < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )
